Estimating Substitution Using Text Embeddings: \\ Evidence from the Film Industry

### 2024-09-22 draft/draft.tex
Include grammar fixes from mom

\author{Gideon Moore}
\title{Estimating Substitution Using Text Embeddings: \\ Evidence from the Movie Industry}

@@

\begin{abstract}
    Using text descriptions of movies in conjunction with weekly box office receipts, I develop a novel model of characteristic-space competition in the movie industry. By exploiting plausably exogeneous variation in movie release windows, I identify the impact of competitor characteristics on movie revenue. As movies become more similar, the impact of competition increases. Due to the film industry's thin profit margins and high fixed costs, replacing a competitor in the 10th percentile of similarity with one in the 90th percentile can reduce profit by as much as 47\%. 
\end{abstract}

@@

Demand estimation is a central pillar of the economist toolkit, informing antitrust regulation, optimal taxation, and trade policy. However, these methods are often limited to markets in which goods have quantifiable characteristics such as storage in the market for hard drives or wattage in the market for light bulbs. Markets with difficult-to-quantify characteristics should not receive less attention simply because they are difficult to quantify; though it is hard to put a number how funny a movie is, we understand that a studio monopolizing ``funny'' films is likely to be bad for consumers. Thus, it is important economists develop tools to estimate demand in markets for goods with difficult-to-quantify characteristics.

Product-level demand systems such as the Almost Ideal Demand System \parencite{deaton1980AER} can estimate rich substitution patterns without any characteristic data by looking only at substitution patterns under variation; that is, the substitution patterns of good $i$ are uninformative about the patterns of good $j$, even if $i$ and $j$ are actually very similar. These methods often require onerous quantities of data on consumer behavior. For example, estimation of AIDS requires sufficient variation to estimate parameters quadratic in the number of goods. A corollary of this is that these methods require goods to have \emph{already entered the market}. If a firm wishes to estimate substitution for a new product, these methods will have little to offer them.

I propose a new method for quantifying product characteristics using text descriptions. By embedding text descriptions of a product in characteristic space, we as econometricians can identify which products are most similar. Thus, we can predict which products compete most closely with others and estimate how the presence of one product impacts the sales of another. Unlike product-based demand estimation methods, this method can inform research on products which have not yet entered the market, and can fit rich substitution patterns while tuning relatively few parameters.

While this method has broad applicability, I use the film industry as a case study. Movies are a natural laboratory for several reasons. First, movies have a broad consumer base, such that readers have an intuitive sense whether my method ``makes sense''--even if you someone hasn't seen \emph{Spider-Man}, they likely have a sense whether it is more similar to \emph{Man of Steel} or \emph{The Notebook}, and can use this lens to judge my results. Second, movie characteristics are often difficult to fit into a traditional econometric framework: while we can easily measure the wattage of a lightbulb, it is much harder to put a number on how funny or violent a movie is--and harder still given the lack of clean data. Finally, the movie industry has specific market characteristics which make demand estimation straightforward conditional on my method. Uniform pricing shuts down the endogeneous pricing channel of competition, and long production pipelines force studios to choose characteristics without perfect knowledge of their competitors.

Using data from BoxOfficeGuru and TheMovieDB, I estimate the impact of competitor characteristics on film revenue for each weekend from 2000-2019. I find substitution patterns which are plausible in both magnitude and direction, and which are robust to the choice of distance metric. As a movie becomes more similar to its competitor, it becomes more and more detrimental to the competitor's revenue. Replacing a competitor in the 10th percentile of similarity with one in the 90th percentile reduces box-office revenue by 3.3\%. Given films' thin profit margins, this is roughly a 47\% decline in profit.

@@

As argued in \textcite{kenter2015P2AICIKM}, bag-of-words and string comparison models struggle when working with short texts. If one paragraph includes the word ``wizardry'' while the other includes the word ``sorcery,'' a bag of words model will not capture the similarity between the two paragraphs. In large enough documents, it is more likely the two words will appear in similar contexts, but this is difficult to rely on when the sample text is only one or two sentences.

When working with short snippets of text like movie descriptions, it is more profitable to pursue \emph{text embedding methods} as pioneered by Google's word2vec architecture \parencite{mikolov2013}. Rather than counting individual words, word2vec identifies similar words using co-occurances in a training sample, then maps these words to a lower-dimensional space. This allows for the identification of similar texts even when they share few or no words, since it is able to identify that synonyms like ``wizardry'' and ``sorcery'' map to the same concept. I use OpenAI's frontier GPT-3 embeddings, which embeds not only words but entire paragraphs in order to capture a broader context.

@@

The most similar paper to mine is \textcite{compiani2023} which uses text characteristics to estimate demand for technology products on Amazon. However, rather than using more traditional logit substitution, my model can estimate substition patterns non-linearly and non-monotonically with distance between two products. I also benchmark my substitution patterns against ``real world'' user preference data from the MovieLens data, confirming my estimates capture actual substitution patterns. Finally, rather than Euclidean distance, my similarity measure is based on the cosine between the embedding vectors; this measure hews more closely to the industry standard, is more efficient to compute at scale, and avoids curse-of-dimensionality issues which often arise when taking Euclidean distances in high-dimensional space.

@@

My closest predecessor studying competition in embedding space is \textcite{magnolfi2022}, who use t-Stochastic Triplet Embeddings to estimate competition in the cereal industry. I take heavy inspiration from their model, however substitute frontier text embedding methods from OpenAI for their triplet-based embeddings. Both they and I follow \textcite{pinkse2002E}, who first estimate product cross-elasticities as a function of distance in characteristic space.

@@

I benefit greatly from research documenting the uniform pricing puzzle \parencite{orbach2007IRLE, gil2009MS, ho2018MR}. This literature extensively documents movie theaters' use of uniform pricing across heterogeneous movies. While each paper suggests potential explanations for this phenomenon, my findings are agnostic to the true cause of uniform pricing; I simply use it as a convenient assumption to estimate demand.

I contribute to the ongoing literature founded by \textcite{prag1994CE} of determinents of box office demand. The authors find well-rated and well-advertised movies generally sell more tickets, and conditional on those two factors other observables are generally insignificant. \textcite{elberse2003MS} model both the supply and the demand for movies, finding that theatres' decision to screen a movie at all is an important factor for revenue. \textcite{devany1996E, devany1997EI, devany1999JCE, devany2004EDC} study several factors in box office revenue, including word-of-mouth, movie retirement, and the film production function. \textcite{ravid2004B} examine the impact of film \emph{content} on performance, finding that violence increases film profitability while sexual content does not.

@@

\section{The Movie Industry}

I use a case study of the film industry to highlight the value of my method. Text is distinctively important for the movie industry because movie characteristics are otherwise difficult to quantify; for example, both \emph{Batman Begins} and \emph{Ant-Man} are big-budget action movies about animal-based superheros, but even a quick glance at the movie descriptions makes clear they are not close substitutes.\footnote{\emph{Batman Begins}: ``Driven by tragedy, billionaire Bruce Wayne dedicates his life to uncovering and defeating the corruption that plagues his home, Gotham City. Unable to work within the system, he instead creates a new identity, a symbol of fear for the criminal underworld.'' \emph{Ant-Man}: ``Armed with the astonishing ability to shrink in scale but increase in strength, master thief Scott Lang must embrace his inner-hero and help his mentor, Doctor Hank Pym, protect the secret behind his spectacular Ant-Man suit from a new generation of towering threats''} This is particularly important given movies are an ``experience good''--people rely on the description to inform their purchasing decision, as most people will not form opinions through repeat purchases. Recognizing this, we expect firms put more effort into the text descriptions of their movies than comparable firms might in more traditional goods markets, as these decisions are distinctively important to customer decisionmaking. Finally, the market for movies has two distinctive features which make it a strong candidate for demand estimation: long production pipelines and a tradition of uniform pricing.

@@

Movie production timelines are both long and secretive. Since studios must purchase a script, hire a cast, and shoot all \emph{before} promoting a movie, it is difficult for a film to change its characteristics in response to its competitors' characteristics. To provide suggestive evidence of this inelasticity, I highlight the existance of ``twin movies:'' pairs of movies released in close proximity with very similar characteristics. The most famous examples, \emph{A Bug's Life} and \emph{Antz}, both released in fall of 1998 and feature early CGI animation of ants rebelling against oppression.

@@

Given filmmakers' expressed distaste for twin movies, why do they continue to exist? Studios are seemingly unable to change their movies' characteristics in response to their competitors' characteristics; at the point DreamWorks learns of \emph{A Bug's Life}, it is too late for them to adjust \emph{Antz} to be more distinct. Thus, the demand estimation coefficients are identified--they do not capture the strategic interplay of positioning and substitutability, but instead the direct impact of competition on film sales.

@@

One threat to this strategy would be if firms lack flexibility on their films' \emph{content}, but can move the film's release date \emph{forward or backward in time}. If I discover I am releasing my children's movie opposite \emph{Frozen}, I may wish to postpone my release date. This would threaten the assumption of exogeneous characteristics among competitors. I think this is unlikely to be a major concern for a couple of reasons. First, there are certain prime release dates which are likely to be more profitable than others; for example, releasing a family movie in November is much better than doing so in January, as you can capture the holiday season. Thus, if a movie targeted a holiday release, moving its release date to avoid competition would potentially be quite costly since there no longer exists consumer demand. Second, the continued existence of twin movies further suggests this is not a margin firms seem to adjust significantly on: if studios could easily move release dates to avoid competition, we would expect to see fewer twin movies than we do.

@@

An ongoing puzzle within the film industry is the existence of a uniform pricing standard: within a timeslot, cinemas generally charge the same price for all movies regardless of excess demand. As a salient example, ``opening night'' showings of movies often sell out, yet theaters do not raise prices. Similarly, movies late in their run often have significant excess capacity, yet theathers do not lower prices.

@@

First, I argue \emph{movies do not compete on price.} Under more flexible pricing, it is possible two movies in close priximity would lower prices to compete for the same audience, raising the quantity sold overall. However, since prices are fixed I assume that consumer decisions are driven entirely by product characteristics and idiosyncratic taste.

@@

Movie consumers face a discrete choice problem: which of the movies on offer this weekend should they see? Let movie $i$'s attractiveness at time $t$ be given by some $\delta_{it}$. Consumers receive idiosyncratic demand shocks for each movie, and choose the movie which maximizes their utility. Due to the uniform pricing scheme discussed above, price \emph{does not} enter into the consumer's decision; choices are driven entirely by movie quality and idiosyncratic taste.

Aggregating over all consumers, we can express the demand for movie $i$ in week $t$ as a function $\phi$ of the movie's appeal $\delta_{it}$ and the appeal of its competitors $\delta_{-it}$. Again due to the uniform pricing scheme, we can use revenues as a proxy for quantity, since quantity is simply a constant scalar transformation of revenue. Finally, demand for movies varies idiosyncratically by weekend; for example, there is much more demand for movies on Christmas Day than on a random weekend in sunny July. Thus, I include a fixed effect $\alpha_t$ to capture week-by-week variation. I can therefore write the demand for movie $i$ in week $t$ as:

@@

A movie's appeal is composed of two parts. First, a fixed effect $\alpha_i$ captures the movie's quality. For example, we would expect \emph{Star Wars} to have a high $\alpha_i$ to capture the fact it is a popular movie in a vacuum separate from its competitors. Second, movies have an age fixed effect $\lambda_{t - r(i)}$, where $r(i)$ represents movie $i$'s release date. As an experience good, demand for a movie is driven heavily by novelty; thus, the same movie will experience much less demand in its second week than in its first. While it is reasonable to expect these $\lambda$ to decrease with time, I do not impose this; instead, verifying this is true will be a test of model fit later on. Finally, I include an idiosyncratic shock $\xi_{it}$ to capture unobserved demand drivers. Thus, the movie's appeal is given by:

@@

How should the demand for movie $i$ depend on its competitors? I put forward three competitor traits which should impact demand for movie $i$:
\begin{itemize}
    \item \emph{Competitor Quality}: If competitor $j$ has a high $\alpha_j$, it should draw more demand away from movie $i$. For example, \emph{The Lord of the Rings} is broadly a ``better'' movie than \emph{Willow,} despite being similar on paper; thus, $i$ would prefer to compete against \emph{Willow} rather than \emph{The Lord of the Rings}.
    \item \emph{Competitor Age}: If competitor $j$ is more recent, it should draw more demand away from movie $i$. That is, $i$ should broadly prefer to release against movies in their fifth week of showing than in their first week.
    \item \emph{Competitor Similarity}: If competitor $j$ is more similar to movie $i$, it should draw away more demand. For example, if $i$ is \emph{Star Wars}, it is much more enthusiastic to release against \emph{The Notebook} than against \emph{Star Trek}.
\end{itemize}

Let $d_{ij}$ be a measure of similarity between movies $i$ and $j$. In this exercise I use the cosine distance of the embedded text descriptions; however, this could be any measure of similarity. There is no reason to expect similarity to enter linearly; instead, I consider some flexible function of distance $f$.

@@

As written, the model above is difficult to estimate. $\alpha_i$ appears not only in its own demand, but in competitors' as well; thus, we cannot use simple demeaning to absorb these fixed effects. Moreover, the $f$ function interacts multiplicatively with the $\delta_{jt}$ terms, making matrix inversion ineffective. With some clever rearranging, however, this model becomes more tractable:\footnote{Gory algebraic detail of this rearrangement is available in appendix \ref{app:rearrangement}.}
\begin{align*}
    \ln(q_{it}) = \alpha_i &+ \sum_{j \neq i} f(d_{ij}) \alpha_j \\
    &+ \lambda_{t - r(i)} \left(1 + \sum_{j \neq i}^{r(j) = r(i)} f(d_{ij})\right) + \sum_{r(k) \neq r(i)} \left(\lambda_{t - r(k)} \left(\sum_{j}^{r(j) = r(k)} f(d_{ij}) \right)\right) \\
    &+ \alpha_t \\
    &+ \xi_{it} + \sum_{j \neq i} f(d_{ij}) \xi_{jt} \qquad \bigg\} \text{ Mean 0 given $f(d_{ij}) \perp \xi_{jt}$}
\end{align*}

Thus, conditional on the values of $f(d_{ij})$, each movie's log quantity is a linear function of it and its competitors' fixed effects and ages plus a fixed effect for the relevant weekend plus mean-zero noise. Thus, the $\lambda$ and $\alpha$ coefficients are estimable using OLS.

As an example, consider what the data matrix will look like in the following market. Movie 1 released this week. Movie 2 released this week as well, while movie 3 released last week. I omit time fixed effects in this example since we focus on a single period; similarly, I omit time subscripts on the quantities. In this case, we will have the following data matrix where values in ``movie'' columns are used to estimate $\alpha$ values while ``age'' columns are used to estimate $\lambda$ values:

@@

\begin{tabular}{cccccc}
    ln(Quantity) & Movie 1 & Movie 2 & Movie 3 & Age 0 & Age 1 \\
    $\ln(q_{1})$ & 1 & $f(d_{12})$ & $f(d_{13})$ & $1 + f(d_{12})$ & $f(d_{13})$ \\
    $\ln(q_{2})$ & $f(d_{12})$ & 1 & $f(d_{23})$ & $1 + f(d_{12})$ & $f(d_{23})$ \\
    $\ln(q_{3})$ & $f(d_{13})$ & $f(d_{23})$ & 1 & $f(d_{13}) + f(d_{23})$ & $1$ 
\end{tabular}

@@

Note the above is all taking $f$ as given. Following \textcite{magnolfi2022}, let $f$ be a cubic polynomial in distance; that is:

@@

Note this optimization is very expensive, as it requires estimating thousands of $\alpha$ coefficients for each interation as well as computing potentially tens of millions of values of $f(d_{ij})$. To speed estimation, I first optimize $\gamma$ using data from 2000 alone (the first ``quality'' year of my data). Conditional on this optimized $\gamma$, I then estimate $\alpha$ and $\lambda$ on all releases from 2000-2019 (omitting 2020 onward due to the COVID-19 pandemic). This allows me to estimate the full model in a reasonable amount of time. In the future, I could potentially run the optimization over the whole dataset using the research computing cluster; however, optimizing on my personal computer is already quite time-consuming even on the small sample.

@@

In some ways this model is a strange fit for ``characteristic-space'' competition, in that it includes movie fixed effects $\alpha_i$. One of the common virtues of using a characteristic-space model is that it allows us to estimate demand for goods which are not yet in the market by arguing based on similar characteristics. However, the econometrician cannot estimate $\alpha_i$ for a movie which has not yet released.

@@

Moreover, the \emph{additive separability} of the $\alpha$ and $\lambda$ terms goes grants certain counterfactuals extra relevance. Note in the model above $\alpha$ and $\lambda$ never interact multiplicatively. Thus, given competitor similarity $d_{ij}$ and ages $\lambda_{t - r(k)}$, it is possible to perform comparative statics agnostic to the $\alpha$ values. For example, I can ask how demand for movie $i$ will compare if it releases in week 1 rather than week 2 conditional on its competitors. Since the terms containing $\alpha_i$ will be identical in these two scenarios, they drop out and leave a concrete estimate of the change in quantity which is not dependent on the $\alpha_i$ values.

@@

I scrape weekend box office receipts from 1997-2024 from BoxOfficeGuru.com, a website maintained by Gitesh Pandya. Pandya is a film consultant specializing in releasing Indian films for the North American market. For each weekend, the website lists the top 10-20 movies by North American box office revenue.

While BoxOfficeGuru is getting its footing in the late 1990s, the data is reported less consistently; thus, I begin my sample in 2000. To avoid the impact of the COVID-19 pandemic, I end my sample in 2019. When Pandya is on vacation the website is not updated, so I have a few missing weekends in my sample interspersed thoughout my period of interest. I drop these weekends from the sample, as I do not have a good way to impute the missing data. This yields a grand total of 901 weekends for analysis.

@@

I retrieve movie characteristics from the API of TheMovieDB, a user-generated database of movies. Specifically, I clean movie names and years retrieved from BoxOfficeGuru and query the search API for the top result released in the relevant year. For each movie, I can thus retrieve the description, genres, and original release language. Example movie descriptions are visible in table \ref{tab:tmdb_desc}.

TMDB also lists the primary language of each movie. I limit only to movies originally released in English to avoid both non-English movie descriptions and differential sales trends among non-English speaking consumers. In total,my sample contains 2,970 movies over the 20 year span matched between BoxOfficeGuru and TheMovieDB.

>>>>

\author{Gideon Moore}
\title{Estimating Substitution Using Text Embeddings: \\ Evidence from the Film Industry}

@@

\begin{abstract}
    Using text descriptions of films in conjunction with weekly box office receipts, I develop a novel model of characteristic-space competition in the film industry. By exploiting plausably exogeneous variation in film release windows, I identify the impact of competitor characteristics on film revenue. As films become more similar, the impact of competition increases. Due to the film industry's thin profit margins and high fixed costs, replacing a competitor in the 10th percentile of similarity with one in the 90th percentile can reduce profit by as much as 47\%. 
\end{abstract}

@@

Demand estimation methods are central pillars of the economist toolkit, informing antitrust regulation, optimal taxation, and trade policy. However, these approaches are often limited to markets in which goods have quantifiable characteristics such as storage in the market for hard drives or wattage in the market for lightbulbs. Markets with difficult-to-quantify characteristics should not receive less attention simply because they are difficult-to-quantify; though it is hard to put a number on how funny a film is, we understand that a studio monopolizing ``funny'' films is likely to be bad for consumers. Thus, it is important economists develop tools to estimate demand in markets where traditional methods struggle to perform well.

Product-level demand systems such as the Almost Ideal Demand System (AIDS) can estimate rich substitution patterns without any characteristic data by looking only at substitution patterns under variation; that is, the substitution patterns of good $i$ are uninformative about the patterns of good $j$, even if $i$ and $j$ are actually very similar \parencite{deaton1980AER}. These methods often require onerous quantities of data on consumer behavior. For example, estimation of AIDS requires sufficient variation to estimate parameters quadratic in the number of goods. A corollary of this is that these methods require goods to have \emph{already entered the market}. If a firm wishes to estimate substitution for a new product, these methods will have little to offer them.

I propose a new method for quantifying product characteristics using text descriptions. By embedding text descriptions of a product in characteristic space, econometricians can identify which products are most similar. Thus, they can predict which products compete most closely with others and estimate how the presence of one product impacts the sales of another. Unlike product-based demand estimation methods, this method can inform research on products which have not yet entered the market, and can fit rich substitution patterns while tuning relatively few parameters.

While this method has broad applicability, I use the film industry as a case study. Film is a natural laboratory for several reasons. First, films have a broad consumer base, such that readers have an intuitive sense whether my method ``makes sense''--even if someone hasn't seen \emph{Spider-Man}, they likely have a sense whether it is more similar to \emph{Man of Steel} or \emph{The Notebook}, and can use this lens to judge my results. Second, film characteristics are often difficult to fit into a traditional econometric framework: while we can easily measure the wattage of a lightbulb, it is much harder to put a number on how funny or violent a film is--and harder still given the lack of clean data. Finally, the film industry has specific market characteristics which make demand estimation using my method straightforward. Uniform pricing shuts down the endogeneous pricing channel of competition, and long production pipelines force studios to choose characteristics without perfect knowledge of their competitors.

Using data from BoxOfficeGuru and TheMovieDB, I estimate the impact of competitor characteristics on film revenue for each weekend from 2000-2019. I find substitution patterns which are plausible in both magnitude and direction, and which are robust to the choice of distance metric. As a film becomes more similar to its competitor, it becomes more and more detrimental to the competitor's revenue. Replacing a competitor in the 10th percentile of similarity with one in the 90th percentile reduces box-office revenue by 3.3\%. Given films' thin profit margins, this is roughly a 47\% decline in profit.

@@

As argued in \textcite{kenter2015P2AICIKM}, bag of words and string comparison models struggle when working with short texts. If one paragraph includes the word ``wizardry'' while another includes the word ``sorcery,'' a bag of words model will not capture the similarity between the two paragraphs. In large enough documents, it is more likely the two words will appear in similar contexts, but this is difficult to rely on when the sample text is only one or two sentences.

When working with short snippets of text like film descriptions, it is more profitable to pursue \emph{text embedding methods} as pioneered by Google's word2vec architecture \parencite{mikolov2013}. Rather than counting individual words, word2vec identifies similar words using co-occurances in a training sample, then maps these words to a lower-dimensional space. This allows for the identification of similar texts even when they share few or no words, since it is able to identify that synonyms like ``wizardry'' and ``sorcery'' map to the same concept. I use OpenAI's frontier GPT-3 embeddings, which embeds not only words but entire paragraphs in order to capture a broader context.

@@

The most similar paper to mine is \textcite{compiani2023} which uses text characteristics to estimate demand for technology products on Amazon. However, rather than using more traditional logit substitution, my model can estimate substitution patterns non-linearly and non-monotonically with distance between two products. I also benchmark my substitution patterns against ``real world'' user preference data from the MovieLens data, confirming my estimates capture actual substitution patterns. Finally, rather than Euclidean distance, my similarity measure is based on the cosine between the embedding vectors; this measure hews more closely to the industry standard, is more efficient to compute at scale, and avoids curse-of-dimensionality issues which often arise when taking Euclidean distances in high-dimensional space.

@@

My closest predecessor studying competition in embedding space is \textcite{magnolfi2024AM}, who use t-Stochastic Triplet Embeddings to estimate competition in the cereal industry. I take heavy inspiration from their model, however substitute frontier text embedding methods from OpenAI for their triplet-based embeddings. Both they and I follow \textcite{pinkse2002E}, who first estimate product cross-elasticities as a function of distance in characteristic space.

@@

I benefit greatly from research documenting the uniform pricing puzzle \parencite{orbach2007IRLE, gil2009MS, ho2018MR}. This literature extensively documents movie theaters' use of uniform pricing across heterogeneous films. While each paper suggests potential explanations for this phenomenon, my findings are agnostic to the true cause of uniform pricing; I simply use it as a convenient assumption to estimate demand.

I contribute to the ongoing literature founded by \textcite{prag1994CE} of determinents of box office demand. The authors find well-rated and well-advertised films generally sell more tickets, and conditional on those two factors other observables are generally insignificant. \textcite{elberse2003MS} model both the supply and the demand for films, finding that theatres' decision to screen a film at all is an important factor for revenue. \textcite{devany1996E, devany1997EI, devany1999JCE, devany2004EDC} study several factors in box office revenue, including word-of-mouth, film retirement, and the film production function. \textcite{ravid2004B} examine the impact of film \emph{content} on performance, finding that violence increases film profitability while sexual content does not.

@@

\section{The Film Industry}

I use a case study of the film industry to highlight the value of my method. Text is distinctively important for the film industry because film characteristics are otherwise difficult to quantify; for example, both \emph{Batman Begins} and \emph{Ant-Man} are big-budget action films about animal-based superheros, but even a quick glance at the descriptions makes clear they are not close substitutes.\footnote{\emph{Batman Begins}: ``Driven by tragedy, billionaire Bruce Wayne dedicates his life to uncovering and defeating the corruption that plagues his home, Gotham City. Unable to work within the system, he instead creates a new identity, a symbol of fear for the criminal underworld.'' \emph{Ant-Man}: ``Armed with the astonishing ability to shrink in scale but increase in strength, master thief Scott Lang must embrace his inner-hero and help his mentor, Doctor Hank Pym, protect the secret behind his spectacular Ant-Man suit from a new generation of towering threats.''} This is particularly important given films are an ``experience good''--people rely on the description to inform their purchasing decision, as most people will not form opinions through repeat purchases. Recognizing this, we expect firms put more effort into the text descriptions of their films than comparable firms might in more traditional goods markets, as these decisions are distinctively important to customer decision-making. Finally, the market for films has two distinctive features which make it a strong candidate for demand estimation: long production pipelines and a tradition of uniform pricing.

@@

Film production timelines are both long and secretive. Since studios must purchase a script, hire a cast, and shoot all \emph{before} promoting a film, it is difficult for a film to change its characteristics in response to its competitors' characteristics. To provide suggestive evidence of this inelasticity, I highlight the existance of ``twin movies'': pairs of films released in close proximity with very similar characteristics. The most famous examples, \emph{A Bug's Life} and \emph{Antz}, both released in fall of 1998 and feature early CGI animation of ants rebelling against oppression.

@@

Given filmmakers' expressed distaste for twin movies, why do they continue to exist? Studios are seemingly unable to change their films' characteristics in response to their competitors' characteristics; at the point DreamWorks learns of \emph{A Bug's Life}, it is too late for them to adjust \emph{Antz} to be more distinct. Thus, the demand estimation coefficients are identified--they do not capture the strategic interplay of positioning and substitutability, but instead the direct impact of competition on film sales.

@@

One threat to this strategy would be if firms lack flexibility on their films' \emph{content}, but can move the film's release date \emph{forward or backward in time}. If I discover I am releasing my children's movie opposite \emph{Frozen}, I may wish to postpone my release date. This would threaten the assumption of exogeneous characteristics among competitors. I think this is unlikely to be a major concern for a couple of reasons. First, there are certain prime release dates which are likely to be more profitable than others; for example, releasing a family film in November is much better than doing so in January, as you can capture the holiday season. Thus, if a film targeted a holiday release, moving its release date to avoid competition would potentially be quite costly since there no longer exists consumer demand. Second, the continued existence of twin movies further suggests this is not a margin firms seem to adjust significantly on: if studios could easily move release dates to avoid competition, we would expect to see fewer twin movies than we do.

@@

An ongoing puzzle within the film industry is the existence of a uniform pricing standard: within a timeslot, cinemas generally charge the same price for all films regardless of excess demand. As a salient example, ``opening night'' showings of films often sell out, yet theaters do not raise prices. Similarly, films late in their run often have significant excess capacity, yet theaters do not lower prices.

@@

First, I argue \emph{films do not compete on price.} Under more flexible pricing, it is possible two films in close proximity would lower prices to compete for the same audience, raising the quantity sold overall. However, since prices are fixed I assume that consumer decisions are driven entirely by product characteristics and idiosyncratic taste.

@@

Consumers face a discrete choice problem: which of the films on offer this weekend should they see? Let film $i$'s attractiveness at time $t$ be given by some $\delta_{it}$. Consumers receive idiosyncratic demand shocks for each film, and choose the option which maximizes their utility. Due to the uniform pricing scheme discussed above, price \emph{does not} enter into the consumer's decision; choices are driven entirely by film quality and idiosyncratic taste.

Aggregating over all consumers, we can express the demand for film $i$ in week $t$ as a function $\phi$ of the film's appeal $\delta_{it}$ and the appeal of its competitors $\delta_{-it}$. Again due to the uniform pricing scheme, we can use revenues as a proxy for quantity, since quantity is simply a constant scalar transformation of revenue. Finally, demand for films varies idiosyncratically by weekend; for example, there is much more demand for films on Christmas Day than on a random weekend in sunny July. Thus, I include a fixed effect $\alpha_t$ to capture week-by-week variation. I can therefore write the demand for film $i$ in week $t$ as:

@@

A film's appeal is composed of two parts. First, a fixed effect $\alpha_i$ captures the film's quality. For example, we would expect \emph{Star Wars} to have a high $\alpha_i$ to capture the fact it is a popular film in a vacuum separate from its competitors. Second, films have an age fixed effect $\lambda_{t - r(i)}$, where $r(i)$ represents film $i$'s release date. As an experience good, demand for a film is driven heavily by novelty; thus, the same film will experience much less demand in its second week than in its first. While it is reasonable to expect these $\lambda$ to decrease with time, I do not impose this; instead, verifying this is true will be a test of model fit later on. Finally, I include an idiosyncratic shock $\xi_{it}$ to capture unobserved demand drivers. Thus, the film's appeal is given by:

@@

How should the demand for film $i$ depend on its competitors? I put forward three competitor traits which should impact demand for film $i$:
\begin{itemize}
    \item \emph{Competitor Quality}: If competitor $j$ has a high $\alpha_j$, it should draw more demand away from film $i$. For example, \emph{The Lord of the Rings} is broadly a ``better'' film than \emph{Willow,} despite being similar on paper; thus, $i$ would prefer to compete against \emph{Willow} rather than \emph{The Lord of the Rings}.
    \item \emph{Competitor Age}: If competitor $j$ is more recent, it should draw more demand away from film $i$. That is, $i$ should broadly prefer to release against films in their fifth week of showing than in their first week.
    \item \emph{Competitor Similarity}: If competitor $j$ is more similar to film $i$, it should draw away more demand. For example, if $i$ is \emph{Star Wars}, it is much more enthusiastic to release against \emph{The Notebook} than against \emph{Star Trek}.
\end{itemize}

Let $d_{ij}$ be a measure of similarity between films $i$ and $j$. In this exercise I use the cosine distance of the embedded text descriptions; however, this could be any measure of similarity. There is no reason to expect similarity to enter linearly; instead, I consider some flexible function of distance $f$.

@@

As written, the model above is difficult to estimate. $\alpha_i$ appears not only in its own demand, but in competitors' as well; thus, we cannot use simple demeaning to absorb these fixed effects. Moreover, the $f$ function interacts multiplicatively with the $\delta_{jt}$ terms, making matrix inversion ineffective. With some rearranging, however, this model becomes more tractable:\footnote{Gory algebraic detail of this rearrangement is available in appendix \ref{app:rearrangement}.}
\begin{align*}
    \ln(q_{it}) = \alpha_i &+ \sum_{j \neq i} f(d_{ij}) \alpha_j \\
    &+ \lambda_{t - r(i)} \left(1 + \sum_{j \neq i}^{r(j) = r(i)} f(d_{ij})\right) + \sum_{r(k) \neq r(i)} \left(\lambda_{t - r(k)} \left(\sum_{j}^{r(j) = r(k)} f(d_{ij}) \right)\right) \\
    &+ \alpha_t \\
    &+ \xi_{it} + \sum_{j \neq i} f(d_{ij}) \xi_{jt} \qquad \bigg\} \text{ Mean 0 given $f(d_{ij}) \perp \xi_{jt}$}
\end{align*}

Thus, conditional on the values of $f(d_{ij})$, each film's log quantity is a linear function of it and its competitors' fixed effects and ages plus a fixed effect for the relevant weekend plus mean-zero noise. Thus, the $\lambda$ and $\alpha$ coefficients are estimable using OLS.

As an example, consider what the data matrix will look like in the following market. Film 1 released this week. Film 2 released this week as well, while film 3 released last week. I omit time fixed effects in this example since we focus on a single period; similarly, I omit time subscripts on the quantities. In this case, we will have the following data matrix where values in ``film'' columns are used to estimate $\alpha$ values while ``age'' columns are used to estimate $\lambda$ values:

@@

\begin{tabular}{cccccc}
    ln(Quantity) & Film 1 & Film 2 & Film 3 & Age 0 & Age 1 \\
    $\ln(q_{1})$ & 1 & $f(d_{12})$ & $f(d_{13})$ & $1 + f(d_{12})$ & $f(d_{13})$ \\
    $\ln(q_{2})$ & $f(d_{12})$ & 1 & $f(d_{23})$ & $1 + f(d_{12})$ & $f(d_{23})$ \\
    $\ln(q_{3})$ & $f(d_{13})$ & $f(d_{23})$ & 1 & $f(d_{13}) + f(d_{23})$ & $1$ 
\end{tabular}

@@

Note all the above is taking $f$ as given. Following \textcite{magnolfi2022}, let $f$ be a cubic polynomial in distance; that is:

@@

Note this optimization is very expensive, as it requires estimating thousands of $\alpha$ coefficients for each interation as well as computing potentially tens of millions of values of $f(d_{ij})$. To speed estimation, I first optimize $\gamma$ using data from 2000 alone (the first ``quality'' year of my data). Conditional on this optimized $\gamma$, I then estimate $\alpha$ and $\lambda$ on all releases from 2000-2019 (omitting 2020 onward due to the negative impact of the COVID-19 pandemic). This allows me to estimate the full model in a reasonable amount of time. In the future, I could potentially run the optimization over the whole dataset using the research computing cluster; however, optimizing on my personal computer is already quite time-consuming even on the small sample.

@@

In some ways this model is a strange fit for ``characteristic-space'' competition, in that it includes film fixed effects $\alpha_i$. One of the common virtues of using a characteristic-space model is that it allows us to estimate demand for goods which are not yet in the market by arguing based on similar characteristics. However, the econometrician cannot estimate $\alpha_i$ for a film which has not yet released.

@@

Moreover, the \emph{additive separability} of the $\alpha$ and $\lambda$ terms grants certain counterfactuals extra relevance. Note in the model above $\alpha$ and $\lambda$ never interact multiplicatively. Thus, given competitor similarity $d_{ij}$ and ages $\lambda_{t - r(k)}$, it is possible to perform comparative statics agnostic to the $\alpha$ values. For example, I can ask how demand for film $i$ will compare if it releases in week 1 rather than week 2 conditional on its competitors. Since the terms containing $\alpha_i$ will be identical in these two scenarios, they drop out and leave a concrete estimate of the change in quantity which is not dependent on the $\alpha_i$ values.

@@

I scrape weekend box office receipts from 1997-2024 from BoxOfficeGuru.com, a website maintained by Gitesh Pandya. Pandya is a film consultant specializing in releasing Indian films for the North American market. For each weekend, the website lists the top 10-20 films by North American box office revenue.

While BoxOfficeGuru was getting its footing in the late 1990s, their reports were less consistent; thus, I begin my sample in 2000. To avoid the negative shock of the COVID-19 pandemic, I end my sample in 2019. When Pandya is on vacation the website is not updated, so I have a few missing weekends in my sample interspersed thoughout my period of interest. I drop these weekends from the sample, as I do not have a good way to impute the missing data. This yields a grand total of 901 weekends for analysis.

@@

I retrieve film characteristics from the API of TheMovieDB, a user-generated database of films. Specifically, I clean film names and years retrieved from BoxOfficeGuru and query the search API for the top result released in the relevant year. For each film, I can thus retrieve the description, genres, and original release language. Example film descriptions are visible in table \ref{tab:tmdb_desc}.

TMDB also lists the primary language of each film. I limit only to films originally released in English to avoid both non-English film descriptions and differential sales trends among non-English speaking consumers. In total, my sample contains 2,970 films over the 20 year span matched between BoxOfficeGuru and TheMovieDB.
